---
output:
  html_document: default
  pdf_document: default
---
Red Wine Quality Exploration by Anupam Chakraborty
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
library(GGally)
library(gridExtra)
library(dplyr)

```

```{r echo=FALSE, Load_the_Data}

# Load the Data
dataRedWine <- read.csv('wineQualityReds.csv')
# Column X contains row number 
dataRedWine$X <- NULL

```

> **Data set: **This tidy data set contains 1,599 red wines with 11 variables on 
the chemical properties of the wine. At least 3 wine experts rated the quality 
of each wine, providing a rating between 0 (very bad) and 10 (very excellent). 

Number of Instances: red wine - 1599  
Number of Attributes: 11 + output attribute

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots_1}

str(dataRedWine)

```


```{r echo=FALSE, Univariate_Plots_2}

summary(dataRedWine)

```

> **Summary of the red wine dataset**: Red wine dataset consists of 12
variables, for 1599 red wine samples

```{r echo=FALSE, Univariate_Function_Definition}

# Function to create Histogram
# Usage: create_histogram('quality', binwidth=1)
create_histogram <- function(variable, binwidth = 0.01) {
  return(ggplot(aes_string(x = variable), data = dataRedWine) + 
           geom_histogram(binwidth = binwidth))
}

# Function to create Histogram after removing outliers
# Usage: create_histogram_without_outlier('alcohol', 
#                    dataRedWine$alcohol, binwidth=0.2)
# Return: It returns a ggplot histogram object
create_histogram_without_outlier <- function(variable, data_var, 
                                             binwidth = 0.01) {
  x1 <- quantile(data_var, 0.01)
  x2 <- quantile(data_var, 0.99)
  return(ggplot(aes_string(x = variable), data = dataRedWine) + 
           ggtitle("Without outlier") +
           geom_histogram(binwidth = binwidth) +
           scale_x_continuous(limits = c(x1, x2)))
}

```



```{r echo=FALSE, Univariate_Plots_3, message=FALSE, warning=FALSE}

create_histogram(variable = 'quality', binwidth=1)

```



```{r echo=FALSE, Univariate_Plots_4}

quality_table <- table(dataRedWine$quality)
quality_table

```
  
The above histogram shows the distribution of red wine quality in the dataset.  
We can see than most wines are of average quality 5, 6.  
Number of wines with quality rating 3 (poor) is: 10  
Number of wines with quality rating 8 (very good): 18  
We will compare wines of quality-3 with quality-8 to understand what makes a 
wine better.  

```{r echo=FALSE, Univariate_Plots_4a}

poorRedWinde <- dataRedWine[dataRedWine$quality<=3, ]
summary(poorRedWinde)

```
  
Summary for wines of quality 3.  
  
```{r echo=FALSE, Univariate_Plots_4b}

goodRedWine <- dataRedWine[dataRedWine$quality >= 8, ]
summary(goodRedWine)

```
  
Summary of wines with quality 8.  
From the summary we notice that there is a difference in alcohol, 
volatile.acidity, citric.acid, chloride between poor and good wines.    

```{r echo=FALSE, Univariate_Plots_5, message=FALSE, warning=FALSE}

p1 <- create_histogram('alcohol', binwidth = 0.2)
p2 <- create_histogram('log10(alcohol)')
p3 <- create_histogram('sqrt(alcohol)', binwidth = 0.02)

grid.arrange(p1, p2, p3, ncol = 1)

```

```{r echo=FALSE, Univariate_Plots_5a, message=FALSE, warning=FALSE}

create_histogram_without_outlier('alcohol', dataRedWine$alcohol, binwidth=0.2)

```

```{r echo=FALSE, Univariate_Plots_6, message=FALSE, warning=FALSE}

summary(dataRedWine$alcohol)

```
  
The above 3 histograms in a grid show distribution of alcohol in the Red Wines 
in different scales.
The 4th histogram plots alcohol removing the outliers (top & bottom 3% data 
values).   
The minimum percentage of alcohol in the sample is 8.40%, the maximum 
percentage is 14.90%.  
50% of the sample has alcohol percentage between 9.50% to 11.10%.  
  
```{r echo=FALSE, Univariate_Plots_7, message=FALSE, warning=FALSE}

create_histogram('pH')
create_histogram_without_outlier('pH', dataRedWine$pH)

```

```{r echo=FALSE, Univariate_Plots_7a, message=FALSE, warning=FALSE}

summary(dataRedWine$pH)

```
  
The above histogram plot distribution of pH in the Red Wine sample.  
In this Red Wine dataset pH values are almost normally distributed.  
Most wines have pH value in the range 3-4.  
  
```{r echo=FALSE, Univariate_Plots_8, message=FALSE, warning=FALSE}
# According documentation most wines are in range 3-4
outsideRangePH <- dataRedWine[dataRedWine$pH < 3 | dataRedWine$pH > 4, ]
table(outsideRangePH$quality)

```
  
Are those wines with pH value out of range poor in quality?  
Wines with pH value out of range (3-4) are distributed over all qualities.  
  
```{r echo=FALSE, Univariate_Plots_9, , message=FALSE, warning=FALSE}

create_histogram('density', binwidth = 0.0005)

```
  
```{r echo=FALSE, Univariate_Plots_9a, , message=FALSE, warning=FALSE}

summary(dataRedWine$density)

```
  
Density of alcohols are normally distributed with mean 0.9967 & median 0.9968.
We know in a Normal distribution mean & median are equal.
  
```{r echo=FALSE, Univariate_Plots_10, message=FALSE, warning=FALSE}

create_histogram('volatile.acidity', binwidth=0.01)
create_histogram_without_outlier('volatile.acidity', 
                                 dataRedWine$volatile.acidity, binwidth=0.01)

```
  
These histograms show the distribution of volatile.acidity in the Red Wines.  
  
```{r echo=FALSE, Univariate_Plots_11, message=FALSE, warning=FALSE}

quantile95 <- quantile(dataRedWine$volatile.acidity, 0.95)
print(quantile95)
# high volatile acidity lead to an unpleasant, vinegar taste
vinegarTasteWine <- dataRedWine[dataRedWine$volatile.acidity >= quantile95, ]
table(vinegarTasteWine$quality)
(table(vinegarTasteWine$quality) / quality_table) *100

```
   
Too high levels of volatile acidity lead to an unpleasant, vinegar taste.   
50% of the wines in quality-3 has vinegar taste.  
30% of the wines in quality-4 has vinegar taste.

```{r echo=FALSE, Univariate_Plots_12, message=FALSE, warning=FALSE}

create_histogram('citric.acid', binwidth = 0.02)
create_histogram_without_outlier('citric.acid', dataRedWine$citric.acid, 
                                 binwidth = 0.02)

```

```{r echo=FALSE, Univariate_Plots_12a, , message=FALSE, warning=FALSE}

summary(dataRedWine$citric.acid)

```  

The above histogram of citric.acid is after removing the outliers.
  
```{r echo=FALSE, Univariate_Plots_13, message=FALSE, warning=FALSE}

create_histogram('fixed.acidity', binwidth = 0.1)
create_histogram('log10(fixed.acidity)', binwidth = 0.025)

```

```{r echo=FALSE, Univariate_Plots_13a, , message=FALSE, warning=FALSE}

summary(dataRedWine$fixed.acidity)

```

The distribution of fixed.acidity is little bit skewed. The distribution of
log(fixed.acidity) is almost normalized.
  
```{r echo=FALSE, Univariate_Plots_14, message=FALSE, warning=FALSE}

create_histogram('residual.sugar', binwidth = 0.10)
create_histogram('log10(residual.sugar)', binwidth= 0.05)

```


```{r echo=FALSE, Univariate_Plots_14a, message=FALSE, warning=FALSE}
summary(dataRedWine$residual.sugar)

```

The distribution of residual.sugar has long tail. In this sample the max value
of residual.sugar is 15.50.  
There is no sweet wine in the sample with greater than 45 grams/liter residual 
sweet.  
Since the value is on permitted range and plotted a histogram with 
log10(residual.sugar).

```{r echo=FALSE, Univariate_Plots_16, message=FALSE, warning=FALSE}

create_histogram('free.sulfur.dioxide', binwidth = 1)
create_histogram('log10(free.sulfur.dioxide)', binwidth = 0.05)

```
```{r echo=FALSE, Univariate_Plots_16b, , message=FALSE, warning=FALSE}

summary(dataRedWine$free.sulfur.dioxide)

```

```{r echo=FALSE, Univariate_Plots_16a, message=FALSE, warning=FALSE}

nrow(dataRedWine[dataRedWine$free.sulfur.dioxide >= 60, ])

```

Plotted histogram of free.sulfur.dioxide. We have seen 2 wines with
very high free.sulfur.dioxide than others.

```{r echo=FALSE, Univariate_Plots_17, message=FALSE, warning=FALSE}

create_histogram('total.sulfur.dioxide', binwidth = 1)
create_histogram('log10(total.sulfur.dioxide)', binwidth = 0.05)

```

```{r echo=FALSE, Univariate_Plots_17b, , message=FALSE, warning=FALSE}

summary(dataRedWine$total.sulfur.dioxide)

```


```{r echo=FALSE, Univariate_Plots_17a, message=FALSE, warning=FALSE}

nrow(dataRedWine[log10(dataRedWine$total.sulfur.dioxide) >= 2.25,])

```

Plotted histogram of log10(total.sulfur.dioxide). We have seen 2 wines with
very high total.sulfur.dioxide than others.

```{r echo=FALSE, Univariate_Plots_18, message=FALSE, warning=FALSE}

create_histogram('sulphates', binwidth = 0.01)
create_histogram('sqrt(sulphates)', binwidth = 0.01)

```

```{r echo=FALSE, Univariate_Plots_18b, , message=FALSE, warning=FALSE}

summary(dataRedWine$sulphates)

```

```{r echo=FALSE, Univariate_Plots_18a, message=FALSE, warning=FALSE}

nrow(dataRedWine[sqrt(dataRedWine$sulphates) >= 1.25,])

```

Sulphates has a skewed distribution. We plotted histogram of sqrt(sulphates) 
and 7 wines have very high sulphates compared to other wines in the sample.

```{r echo=FALSE, Univariate_Plots_19, message=FALSE, warning=FALSE}

create_histogram('chlorides', binwidth = 0.005)
create_histogram('log10(chlorides)', binwidth = 0.01)

```
```{r echo=FALSE, Univariate_Plots_19b, , message=FALSE, warning=FALSE}

summary(dataRedWine$chlorides)

```

log10(chlorides) is almost normally distributed.

# Univariate Analysis

### What is the structure of your dataset?

Data set contains 1,599 red wines data on 12 attributes. 11 variables 
like alcohol, pH, density, acidity, residual.sigar etc. on the chemical 
properties of the wine. The quality attribute is provided by wine experts in a 
scale of 0 to 10. In this dataset wines have quality between 3 and 8. Following 
is the distribution of wine quality in the dataset. There are only 10 poor wine 
(quality-3) & 18 good wine (quality-8).

  3   4   5   6   7   8  
 10  53 681 638 199  18 

Poor quality & good quality wines differs in the percentage of alcohol & 
amount of volatile.acidity, citric.acid, chloride.    

### What is/are the main feature(s) of interest in your dataset?
The main features of this dataset is quality, alcohol, volatile.acidity. I want 
to investigate how to classify wines' quality based on alcohol, 
volatile.acidityand other features.

### What other features in the dataset do you think will help support your
The other features that could influence the quality of wine are sulphate,
citic.acid, total.sulfur.dioxide, density. 

### Did you create any new variables from existing variables in the dataset?
I haven't created any new variable. The quality can be used as factor variable
using as.factor(quality).

### Of the features you investigated, were there any unusual distributions?
Plotted histogram of alcohol with actual, log10 & square root.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots_1, fig.width=9, fig.height=6}

ggpairs(dataRedWine)

```
```{r echo=FALSE, Bivariate_Plots_1a, fig.width=9, fig.height=6}

ggcorr(dataRedWine, label = TRUE, label_size = 3, label_round = 2, 
       label_alpha = TRUE)

```

```{r echo=FALSE, Bivariate_Function_Definition}

# Draws Box Plot, Where x is 'quality'
# Usage: create_boxplot('alcohol')
# Return: It returns a ggplot boxplot object
create_boxplot <- function(variable) {
  return(ggplot(aes_string(x = 'as.factor(quality)', y = variable), 
                data = dataRedWine) + 
           geom_boxplot(alpha = .5,color = 'blue') +
           geom_jitter( alpha = .3)  +
             stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4))
}

```


```{r echo=FALSE, Bivariate_Plots_2}

create_boxplot('alcohol')
  
```


```{r echo=FALSE, Bivariate_Plots_3}

cor.test(dataRedWine$quality, dataRedWine$alcohol, method='pearson')

```

Creating box plot for categorcial variable (quality) and continuous variable
(alcohol percentage). As quality of Red wine is highly correlated with alcohol 
percentage (correlation coefficient: 0.47) compared to any other chemical 
properties of Red wine.

```{r echo=FALSE, Bivariate_Plots_4}

by(dataRedWine$alcohol ,dataRedWine$quality, summary)

```

Above scatter plot, Pearson's correlation, box plot & summary statistics
show that mean & median of alcohol content in better quality of wine 
is more than inferior quality wine.  

```{r echo=FALSE, Bivariate_Plots_6}

create_boxplot('pH')

```

```{r echo=FALSE, Bivariate_Plots_6A}

cor.test(dataRedWine$quality, dataRedWine$pH, method='pearson')

```

Above plot of pH mean, median Vs quality shows pH doesn't influence quality of 
wine.


```{r echo=FALSE, Bivariate_Plots_7}

create_boxplot('volatile.acidity')

```

Creating box plot as quality of Red wine has second highest correlation with 
volatile.acidity (correlation coefficient: -0.39) compared to other chemical 
properties of Red wine.

```{r echo=FALSE, Bivariate_Plots_7a}

cor.test(dataRedWine$quality, dataRedWine$volatile.acidity, method = 'pearson')

```

Above scatter plot, Pearson's correlation, mean of each quality show
that volatile acidity is less in better quality of wine than inferior quality 
wine.  

```{r echo=FALSE, Bivariate_Plots_12}

create_boxplot('sulphates')

```
```{r echo=FALSE, Bivariate_Plots_12a}

cor.test(dataRedWine$quality, dataRedWine$sulphates, method = 'pearson')

```
Creating a box plot as quality has third highest correlation (0.25) with 
sulphates. Above plot & Pearson's correlation shows that sulphate is more in 
better quality wine. 


```{r echo=FALSE, Bivariate_Plots_10}

create_boxplot('citric.acid')

```
```{r echo=FALSE, Bivariate_Plots_10a}

cor.test(dataRedWine$quality, dataRedWine$citric.acid, method = 'pearson')

```  
Creating a box plot as quality has fourth highest correlation (0.23) with 
citric.acid. Above box shows that mean of citric acid is higher in better 
quality of wine than inferior quality wine. 

```{r echo=FALSE, Bivariate_Plots_11}

create_boxplot('density')

```
```{r echo=FALSE, Bivariate_Plots_11a}

cor.test(dataRedWine$quality, dataRedWine$density, method = 'pearson')

```
Creating a box plot as quality and density has a correlation coefficient -0.17.
Above plot shows that density is less in better quality wine. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation  
We have find the following relationships between the chemical composition of 
wine and its quality:  
1) Better quality red wines have more percentage of alcohol.  
2) Better quality red wines have less volatile acidity but more citric acid.  
3) Density of better quality of wine is less.  
4) Better quality wine has more sulphates.  

### Did you observe any interesting relationships between the other features
The relationship between other variables  
1) As the alcohol content increases, density decreases. As better wines have
more alcohol percentage that causes lesser density.  
2) Total Sulfur dioxide is proportional to free sulfur dioxide in wine.  
3) citric acid in proportional to the fixed.acidity in wine.

### What was the strongest relationship you found?
The better quality wines have more percentage of alcohol and thus have less
density.

# Multivariate Plots Section

```{r echo=FALSE, Trivariate_Function_Definition}

# Draws scatter plot after removing outliers
# Usage: create_scatterplot('alcohol', 'density', dataRedWine$alcohol, 
# dataRedWine$density)
# Return: It returns a ggplot scatterplot object. 
# Each data point is colored in the scatter plot according its quality.
# A regression line added for each category to depict the separation.
create_scatterplot <- function(variable_x, variable_y, data_var_x, 
                               data_var_y) {
  x1 <- quantile(data_var_x, 0.01)
  x2 <- quantile(data_var_x, 0.99)
  y1 <- quantile(data_var_y, 0.01)
  y2 <- quantile(data_var_y, 0.99)
  
  return(ggplot(aes_string(x = variable_x, 
           y = variable_y, color = 'as.factor(quality)'), 
           data = dataRedWine) +
      xlim(x1, x2) +
      ylim(y1, y2) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE, size=1)  +
      scale_color_brewer(type='seq',
           guide=guide_legend(title='Quality')))
}

# Draws scatter plot after removing outliers
# Usage: create_scatterplot('alcohol', 'density', dataRedWine$alcohol, 
# dataRedWine$density)
# Return: It returns a ggplot scatterplot object
create_scatterplot_colour_quality <- function(variable_x, variable_y, 
                                              data_var_x, data_var_y) {
  x1 <- quantile(data_var_x, 0.01)
  x2 <- quantile(data_var_x, 0.99)
  y1 <- quantile(data_var_y, 0.01)
  y2 <- quantile(data_var_y, 0.99)
  
  return(ggplot(aes_string(x = variable_x, y = variable_y), 
         data = dataRedWine) +
    xlim(x1, x2) +
    ylim(y1, y2) +
    geom_point(alpha = 1/2, position = position_jitter(h = 0),
               aes(color = as.factor(quality))) + 
    scale_color_brewer(type='div',
                   guide=guide_legend(title='Quality')) +
    theme_dark())
}

```


```{r echo=FALSE, Bivariate_Plots_14, message=FALSE, warning=FALSE}

create_scatterplot('alcohol', 'density', dataRedWine$alcohol, 
                   dataRedWine$density)
cor.test(dataRedWine$alcohol, dataRedWine$density, method = 'pearson')

```
Created a scatter plot between alcohol & density as they have very high 
correlation of -0.49.

```{r echo=FALSE, Bivariate_Plots_15, message=FALSE, warning=FALSE}


create_scatterplot('total.sulfur.dioxide','free.sulfur.dioxide', 
                   dataRedWine$total.sulfur.dioxide, 
                   dataRedWine$free.sulfur.dioxide)

cor.test(dataRedWine$total.sulfur.dioxide, dataRedWine$free.sulfur.dioxide, 
         method = 'pearson')

```

Created a scatter plot between total.sulfur.dioxide & free.sulfur.dioxide as 
they have very high correlation of 0.67.

```{r echo=FALSE, Bivariate_Plots_16, message=FALSE, warning=FALSE}

create_scatterplot('fixed.acidity', 'citric.acid', 
                   dataRedWine$fixed.acidity, dataRedWine$citric.acid)

cor.test(dataRedWine$fixed.acidity, dataRedWine$citric.acid, 
         method = 'pearson')

```

Created a scatter plot between fixed.acidity & citric.acid as 
they have very high correlation of 0.67.

```{r echo=FALSE, Multivariate_Plots, message=FALSE, warning=FALSE}
create_scatterplot_colour_quality('alcohol', 'volatile.acidity', 
                dataRedWine$alcohol, dataRedWine$volatile.acidity) 

```

In the above plot of alcohol Vs volatile acidity:  
- top left quadrant is for wines with less alcohol & more volatile acidity, 
most of the wines are of bad quality here.  
- bottom right quadrant is for wines with more alcohol & less volatile acidity, 
most of the wines are of good quality here.  


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
We have seen in bivariate distribution that better quality of wines have 
more alcohols and less volatile acidity. We wanted to plot wine quality in a 
scatter plot of alcohol Vs volatile.acidity.  
- top left quadrant is for wines with less alcohol & more volatile acidity, 
most of the wines are of bad quality here.  
- bottom right quadrant is for wines with more alcohol & less volatile acidity, 
most of the wines are of good quality here.

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

  create_boxplot('alcohol') +
  ggtitle("Red Wine Quality Vs Alcohol Percentage") +
  xlab("Wine quality (0 to 10)") +
  ylab("Percentage of alcohol")
  
```

### Description One
Above scatter plot shows that mean & median of alcohol content for each quality 
of wines. It is evident from the plot that better quality wines have more 
alcohol percentageis than inferior quality wines.  

### Plot Two

```{r echo=FALSE, Plot_Two}

  create_boxplot('volatile.acidity') +
  ggtitle("Red Wine Quality Vs Volatile Acidity") +
  xlab("Wine quality (0 to 10)") +
  ylab("Volatile acidity (g / dm^3)")

```

### Description Two
Above scatter plot shows mean & median of volatile acidity for each quality.
The volatile acidity is less in better quality of wine than inferior quality 
wine.  

### Plot Three

```{r echo=FALSE, Plot_Three, message=FALSE, warning=FALSE}
create_scatterplot_colour_quality('alcohol', 'volatile.acidity', 
                dataRedWine$alcohol, dataRedWine$volatile.acidity) +
  ggtitle("Alcohol Vs Volatile Acidity to Classify Wine Quality") +
  xlab("Percentage of alcohol") +
  ylab("Volatile acidity (g / dm^3)")

```

### Description Three
We have seen in bivariate distribution that better quality of wines have  
more alcohols and less volatile acidity. We wanted to plot wine quality in a
scatter plot of alcohol Vs volatile.acidity.
- top left quadrant is for wines with less alcohol & more volatile acidity, 
most of the wines are of bad quality here.
- bottom right quadrant is for wines with more alcohol & less volatile acidity, 
most of the wines are of good quality here.

------

# Reflection
The Red Wine data set contains information of 1599 wines across twelve 
attributes. I started by understanding the individual variables in the 
data set, and then I explored interesting questions and leads as I continued 
to make observations on plots. Eventually, I explored the quality of wine 
across many variables and tried to create a classifier.

There was a clear trend between the alcohol percentage of a wine and its 
quality. Another trend observed between residual acidity and wine quality. 
I was surprised that pH did not have a strong positive correlation with quality.

The biggest challenge I faced in analysing Red Wine dataset is to find outliers. 
Outlier can be  an accurate data of an extreme case. I though outliers will be 
helpful to classify wines as bad or good. So I carefully analyse how outliers 
are distributed to different qualities. Following is the example of distribution 
of wines with pH in unusual range less than 3 or greater than 4.
4  5  6  7  8 
 1 13 12  3  2 


I tried to create the classifier based on the alcohol percentage and volatile 
acidity in multivariate plot. A wine that is plotted in the top left quadrant 
most probably is not a good quality one. A wine that is plotted in the bottom
right quadrant is most probably a good quality wine with more alcohol & less
volatile acidity.

One area of improvement is R functions that I have written. Currently I am
passing very similar parameters to call functions, like:
create_scatterplot_colour_quality('alcohol', 'volatile.acidity', 
                dataRedWine$alcohol, dataRedWine$volatile.acidity)
I want to reduce the function call to:
                create_scatterplot_colour_quality('alcohol', 'volatile.acidity')
